The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women

### 2025-04-03 code/analysis-code/Statistical_Analysis.qmd
update

```{r}
# **1. Load Required Libraries**
# Essential libraries for data handling, visualization, and analysis

library(here)             # Manage file paths
library(dplyr)            # Data manipulation
library(ggplot2)          # Visualization
library(tidyverse)        # Data wrangling and cleaning
library(janitor)          # Cleaning column names
library(skimr)            # Quick data summary
library(lubridate)        # Working with dates
library(readr)            # Reading CSV files
library(gtsummary)        # Creating summary tables
library(gt)               # Formatting tables
library(knitr)            # Table formatting
library(kableExtra)       # Additional table customization
library(ggpubr)           # Publication-ready plots
library(Amelia)           # Missing data visualization
library(forcats)          # Working with categorical variables
library(pwr)              # Power analysis
library(DiagrammeR)       # Flowcharts/Diagrams
library(survival)         # Survival analysis
library(survminer)        # Survival plots
```

```{r}
# Load the here package
library(here)

# Construct the relative file path and import the CSV data
promo_data <- read.csv(here("data", "raw-data", "PROMO_Data.csv"))

# View the first few rows of the data
head(promo_data)

@@

## Part 2: Data Loading, Cleaning, Wrangling, and Exploratory Analysis

# Data Cleaning & Transformation
```{r}
# Display the structure of the data
str(promo_data)

# Print the column names
colnames(promo_data)

# Display the first few rows of the data
head(promo_data)

# Provide summary statistics for each variable
summary(promo_data)

@@

```{r}
# Import the CSV data using here
promo_data <- read.csv(here("data", "raw-data", "PROMO_Data.csv"))

# Clean column names (all lower-case with underscores)
promo_data_clean <- promo_data %>% 
  clean_names()

# Convert date columns to Date objects
promo_data_clean <- promo_data_clean %>%
  mutate(
    enrollment_date = as.Date(enrollment_date, format = "%Y-%m-%d"),
    withdrawal_date = as.Date(withdrawal_date, format = "%Y-%m-%d"),
    child_withdrawal_date = as.Date(child_withdrawal_date, format = "%Y-%m-%d")
  )

@@

# View a summary of the cleaned data
summary(promo_data_clean)

```{r}
# Ensure the 'data/clean' directory exists
if (!dir.exists(here("data", "clean"))) {
  dir.create(here("data", "clean"), recursive = TRUE)
}

@@

# Exploratory Data Analysis
```{r}
# Create a histogram of total malaria episodes
ggplot(data = promo_data_clean, aes(x = total_malaria_episodes)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Total Malaria Episodes",
    x = "Total Malaria Episodes",
    y = "Count"
  )

@@

@@

# Create a scatter plot with regression lines, faceted by study arm
ggplot(promo_data_clean, aes(x = total_malaria_episodes, y = child_1_birth_weight_kg)) +
  geom_point(alpha = 0.7, color = "darkblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~ study_arm) +
  labs(
    title = "Malaria Episodes vs. Birth Weight by Study Arm",
    x = "Total Malaria Episodes",
    y = "Child Birth Weight (kg)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

@@

## 1. Univariate Analysis

# Histogram: Distribution of Age at Enrollment
ggplot(promo_data_clean, aes(x = age_at_enrollment_years)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Age at Enrollment",
       x = "Age (years)", y = "Count")

# Histogram: Distribution of Gestational Age at Enrollment
# Adjust the variable name if necessary after cleaning (e.g., gestational_age_at_enrollment_weeks)
ggplot(promo_data_clean, aes(x = gestational_age_at_enrollment_weeks)) +
  geom_histogram(binwidth = 2, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Gestational Age at Enrollment",
       x = "Gestational Age (weeks)", y = "Count")

# Histogram: Distribution of Total Malaria Episodes
ggplot(promo_data_clean, aes(x = total_malaria_episodes)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Total Malaria Episodes",
       x = "Total Malaria Episodes", y = "Count")

# Bar Plot: Counts of Study Arm
ggplot(promo_data_clean, aes(x = study_arm)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Counts by Study Arm", x = "Study Arm", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Bar Plot: Counts of Adverse Outcomes (assumed to be coded as 0/1)
ggplot(promo_data_clean, aes(x = factor(adverse_outcome))) +
  geom_bar(fill = "salmon") +
  labs(title = "Counts of Adverse Outcomes",
       x = "Adverse Outcome (0 = No, 1 = Yes)", y = "Count")

## 2. Bivariate Analysis

# Boxplot: Total Malaria Episodes by Adverse Outcome
ggplot(promo_data_clean, aes(x = factor(adverse_outcome), y = total_malaria_episodes)) +
  geom_boxplot(fill = "lightgray") +
  labs(title = "Malaria Episodes by Adverse Outcome",
       x = "Adverse Outcome (0 = No, 1 = Yes)",
       y = "Total Malaria Episodes")

# Boxplot: Total Malaria Episodes by Study Arm
ggplot(promo_data_clean, aes(x = study_arm, y = total_malaria_episodes)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Malaria Episodes by Study Arm",
       x = "Study Arm",
       y = "Total Malaria Episodes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Contingency Table: Study Arm vs. Adverse Outcome
study_adverse_table <- table(promo_data_clean$study_arm, promo_data_clean$adverse_outcome)
print(study_adverse_table)

# Stacked Bar Plot: Proportion of Adverse Outcomes by Study Arm
ggplot(promo_data_clean, aes(x = study_arm, fill = factor(adverse_outcome))) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Adverse Outcomes by Study Arm",
       x = "Study Arm", y = "Proportion", fill = "Adverse Outcome") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

## 3. Multivariate/Correlation Analysis

# Select continuous variables of interest
continuous_vars <- promo_data_clean %>%
  select(age_at_enrollment_years, gestational_age_at_enrollment_weeks, 
         total_malaria_episodes, child_1_birth_weight_kg)

# Compute correlation matrix (using complete observations)
cor_matrix <- cor(continuous_vars, use = "complete.obs")
print(cor_matrix)

# Visualize the correlation matrix
corrplot(cor_matrix, method = "circle", type = "upper")

## 4. Visual Exploration with Scatter Plots

# Scatter Plot: Total Malaria Episodes vs. Child Birth Weight, with regression line and faceting by Study Arm
ggplot(promo_data_clean, aes(x = total_malaria_episodes, y = child_1_birth_weight_kg)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~ study_arm) +
  labs(title = "Malaria Episodes vs. Birth Weight by Study Arm",
       x = "Total Malaria Episodes",
       y = "Child Birth Weight (kg)")

@@

```{r}
# Install the tableone package if not already installed
if (!requireNamespace("tableone", quietly = TRUE)) {
  install.packages("tableone")
}

library(tableone)
library(here)

# Define the variables for the demographics and baseline characteristics table.
vars <- c("age_at_enrollment_years", 
          "gestational_age_at_enrollment_weeks", 
          "education_level", 
          "gravidity", 
          "parity")

strata <- "study_arm"

# Create the TableOne object
table1 <- CreateTableOne(vars = vars, strata = strata, data = promo_data_clean, factorVars = c("education_level", "study_arm"))

# Print the table with all levels displayed
print(table1, showAllLevels = TRUE)

@@

```{r}
# Create table data using tribble.
# Each row must have exactly 4 cells.
table_data <- tribble(
  ~Characteristic,                         ~Subcategory,                                         ~`Monthly SP (n=391)`,                                         ~`Monthly DP (n=391)`,
  # Age at Enrollment
  "Age, years",                            "",                                                   "23 (19–27)",                                                  "23 (19–27)",
  # Gestational Age at Enrollment overall and subcategories
  "Gestational Age at Enrollment (weeks)", "",                                                   "15.4 (13.3–17.6)",                                              "15.0 (13.4–17.1)",
  "",                                      "12–16",                                              "66 (17%)",                                                    "62 (15.8%)",
  "",                                      "17–20",                                              "164 (42.2%)",                                                 "197 (50.1%)",
  "",                                      "21–24",                                              "133 (34.2%)",                                                 "118 (30%)",
  # Gravidity overall and subcategories
  "Gravidity",                             "",                                                   "1.2 (0.9–1.7)",                                                 "1.1 (0.8–1.6)",
  "",                                      "1",                                                  "102 (26%)",                                                   "95 (24%)",
  "",                                      "2–3",                                                "180 (46%)",                                                   "190 (49%)",
  "",                                      "≥4",                                                 "109 (28%)",                                                   "106 (27%)",
  # Parity overall and subcategories
  "Parity",                                "",                                                   "2.43 (1.67)",                                                   "2.43 (1.65)",
  "",                                      "0",                                                  "140 (36%)",                                                   "145 (37%)",
  "",                                      "1–2",                                                "170 (44%)",                                                   "165 (42%)",
  "",                                      "≥3",                                                 "81 (20%)",                                                    "81 (21%)",
  # Total Malaria Episodes overall
  "Total Malaria Episodes (mean (SD))",    "",                                                   "1.06 (1.03)",                                                 "0.98 (0.95)",
  # Maternal Nutritional Status/BMI overall
  "Maternal Nutritional Status/BMI (mean (SD))", "",                                                "22.5 (3.2)",                                                  "22.8 (3.1)",
  # History of Previous Malaria Episodes overall
  "History of Previous Malaria Episodes (%)", "",                                                 "40 (10%)",                                                    "35 (9%)",
  # Maternal Education Level overall and subcategories
  "Maternal Education Level (%)",          "",                                                   "",                                                              "",
  "",                                      "A and O Level",                                      "194 (49.4%)",                                                 "187 (48.1%)",
  "",                                      "Primary",                                            "68 (17.3%)",                                                  "71 (18.3%)",
  "",                                      "Tertiary",                                           "65 (16.5%)",                                                  "61 (15.7%)",
  "",                                      "University",                                         "66 (16.8%)",                                                  "70 (18.0%)"
)

# Render the table using kableExtra
kbl(
  table_data,
  col.names = c("Characteristic", "Subcategory", "Monthly SP (n=391)", "Monthly DP (n=391)"),
  booktabs = TRUE,
  align = "lccc"
) %>%
  add_header_above(c("N = 782" = 4)) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  # Merge sub-rows for Gestational Age at Enrollment:
  # Overall row is row 2; subcategory rows are rows 3-5
  pack_rows("Gestational Age at Enrollment (weeks)", start_row = 2, end_row = 5, indent = TRUE) %>%
  # Merge sub-rows for Gravidity:
  # Overall row is row 6; subcategory rows are rows 7-9
  pack_rows("Gravidity", start_row = 6, end_row = 9, indent = TRUE) %>%
  # Merge sub-rows for Parity:
  # Overall row is row 10; subcategory rows are rows 11-13
  pack_rows("Parity", start_row = 10, end_row = 13, indent = TRUE) %>%
  # Merge sub-rows for Maternal Education Level:
  # Overall row is row 17; subcategory rows are rows 18-21
  pack_rows("Maternal Education Level (%)", start_row = 17, end_row = 21, indent = TRUE)

@@

```{r}
table(promo_data_clean$stillbirth_bin, useNA = "always")
table(promo_data_clean$stillbirth_bin, useNA = "always")
table(promo_data_clean$stillbirth)

@@

## Results

```{r}
# --- Load libraries ---
library(dplyr)
library(broom)
library(logistf)
library(knitr)
library(kableExtra)
library(tibble)

# --- Data preparation ---
promo_data_clean <- promo_data_clean %>%
  mutate(
    preterm_birth = ifelse(gestational_age_at_delivery_weeks < 37, 1, 0),
    stillbirth_bin = ifelse(stillbirth == "Yes", 1, 0)
  )

# --- Parsimonious Logistic Regression for Preterm Birth ---
model_preterm_pars <- glm(
  preterm_birth ~ total_malaria_episodes_during_pregnancy +
    age_at_enrollment_years + gravidity,
  data = promo_data_clean,
  family = binomial(link = "logit")
)

>>>>

```{r}
# Load Required Libraries
# Essential libraries for data handling, visualization, and analysis

library(here)             # Manage file paths
library(dplyr)            # Data manipulation
library(tidyverse)        # Data wrangling and visualization
library(janitor)          # Clean column names
library(skimr)            # Quick data summary
library(lubridate)        # Working with dates
library(readr)            # Reading CSV files
library(ggplot2)          # Visualization
library(ggpubr)           # Publication-ready plots
library(forcats)          # Categorical variables handling

# Table and reporting packages
library(gtsummary)        # Summary tables
library(gt)               # Table formatting
library(knitr)            # Table rendering
library(kableExtra)       # Table customization

# Analysis-specific packages
library(Amelia)           # Missing data visualization
library(pwr)              # Power analysis
library(DiagrammeR)       # Diagrams and flowcharts
library(survival)         # Survival analysis
library(survminer)        # Survival plots
library(ggeffects)        # Marginal effects for regression models
library(broom)            # Tidy model outputs
library(rsample)          # Resampling and splitting datasets

@@

```{r}
# Data Import & Initial Inspection 
# load the Dataset
PROMO_Data <- read_csv(here("data", "raw-data", "PROMO_Data.csv"))

promo_data_clean <- read.csv(here("data", "clean", "PROMO_Data_clean.csv"))
```

```{r}
names(promo_data_clean)
```

*Table 1*

```{r}
# Recode variables for baseline table presentation:
promo_data_clean <- promo_data_clean %>%
  mutate(
    # Create an age group variable (for potential subgroup analyses)
    age_group = ifelse(age_at_enrollment_years < 25, "Young", "Older"),

    # Recode Gravidity into categories: "1", "2–3", "≥4" (ordered chronologically)
    Gravidity_cat = case_when(
      gravidity == 1 ~ "1",
      gravidity %in% c(2, 3) ~ "2–3",
      gravidity >= 4 ~ "≥4"
    ),
    Gravidity_cat = factor(Gravidity_cat, levels = c("1", "2–3", "≥4")),

    # Recode Parity into categories: "0", "1–2", "≥3" (ordered chronologically)
    Parity_cat = case_when(
      parity == 0 ~ "0",
      parity %in% c(1, 2) ~ "1–2",
      parity >= 3 ~ "≥3"
    ),
    Parity_cat = factor(Parity_cat, levels = c("0", "1–2", "≥3")),

    # Recode Total Malaria Episodes into categories:
    # Combine 0 and 1 episodes as "1", 2-3 as "2–3", and 4 or more as "≥4"
    MalariaEpisodes_cat = case_when(
      total_malaria_episodes %in% c(0, 1) ~ "1",
      total_malaria_episodes %in% c(2, 3) ~ "2–3",
      total_malaria_episodes >= 4 ~ "≥4"
    ),
    MalariaEpisodes_cat = factor(MalariaEpisodes_cat, levels = c("1", "2–3", "≥4")),

    # Recode Total Malaria Episodes During Pregnancy similarly:
    MalariaEpisodesPreg_cat = case_when(
      total_malaria_episodes_during_pregnancy %in% c(0, 1) ~ "1",
      total_malaria_episodes_during_pregnancy %in% c(2, 3) ~ "2–3",
      total_malaria_episodes_during_pregnancy >= 4 ~ "≥4"
    ),
    MalariaEpisodesPreg_cat = factor(MalariaEpisodesPreg_cat, levels = c("1", "2–3", "≥4")),

    # Recode Preterm Births Count into categories:
    # Combine 0 and 1 as "1", and 2 as "2"
    PretermBirths_cat = case_when(
      preterm_births_count %in% c(0, 1) ~ "1",
      preterm_births_count == 2 ~ "2"
    ),
    PretermBirths_cat = factor(PretermBirths_cat, levels = c("1", "2"))
  )

# Create Table X: Baseline Characteristics Stratified by IPTp Treatment Arm
baseline_table_treatment <- promo_data_clean %>%
  dplyr::select(
    study_arm,
    `Age (years)` = age_at_enrollment_years,
    `Gestational Age (weeks)` = gestational_age_at_enrollment_weeks,
    `Maternal Education Level` = education_level,
    Gravidity = Gravidity_cat,
    Parity = Parity_cat,
    `Total Malaria Episodes` = MalariaEpisodes_cat,
    `Total Malaria Episodes During Pregnancy` = MalariaEpisodesPreg_cat,
    `Malaria Infection Rate During Pregnancy` = malaria_infection_rate_during_pregnancy,
    `Placental Malaria (Rogerson Criteria)` = placental_malaria_by_rogerson_criteria,
    `Preterm Births Count` = PretermBirths_cat,
    `Stillbirth bin` = stillbirth_bin,
    `Birthweight` = birth_weight
  ) %>%
  tbl_summary(
    by = study_arm,
    missing = "no",
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    )
  ) %>%
  modify_header(label = "") %>%
  modify_caption("**Table X: Baseline Characteristics by IPTp Treatment Arm**")

# Print Table X
baseline_table_treatment

@@

*Table 2: Outcome Measures and Malaria Exposure Variables Stratified by IPTp Regimen*

```{r}
# Load necessary libraries
library(dplyr)
library(gtsummary)
library(here)

# Read in the cleaned data (if not already read)
promo_data_clean <- read.csv(here("data", "clean", "PROMO_Data_clean.csv"))

# Recode variables as needed for outcomes analysis:
promo_data_clean <- promo_data_clean %>%
  mutate(
    # Create a composite adverse outcome variable: 1 if any of preterm birth, stillbirth, or low birth weight (<2.5 kg) occurs
    low_birth_weight = ifelse(birth_weight < 2.5, 1, 0),
    adverse_birth_outcome = ifelse(preterm_births_count > 0 | stillbirth_bin == 1 | low_birth_weight == 1, 1, 0),

    # Recode Preterm Births Count as before (0/1 -> "1", 2 -> "2")
    PretermBirths_cat = case_when(
      preterm_births_count %in% c(0, 1) ~ "1",
      preterm_births_count == 2 ~ "2"
    ),
    PretermBirths_cat = factor(PretermBirths_cat, levels = c("1", "2"))
  )

# Create Table Z: Outcome Variables by IPTp Treatment Arm with p-values
table_outcomes <- promo_data_clean %>%
  dplyr::select(
    study_arm,
    `Malaria Infection Rate During Pregnancy` = malaria_infection_rate_during_pregnancy,
    `Placental Malaria (Rogerson Criteria)` = placental_malaria_by_rogerson_criteria,
    `Preterm Births Count` = PretermBirths_cat,
    `Stillbirth bin` = stillbirth_bin,
    `Birthweight` = birth_weight,
    `Composite Adverse Outcome` = adverse_birth_outcome
  ) %>%
  tbl_summary(
    by = study_arm,
    missing = "no",
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    )
  ) %>%
  add_p() %>%
  modify_header(label = "") %>%
  modify_caption("")

# Print Table Z
table_outcomes

# Visualization

*Figure 1: Histogram of Total Malaria Episodes by IPTp Treatment Arm*

```{r}
library(ggplot2)

ggplot(promo_data_clean, aes(x = total_malaria_episodes, fill = study_arm)) +
  geom_histogram(binwidth = 1, alpha = 0.6, position = "dodge") +
  labs(title = "Distribution of Total Malaria Episodes by IPTp Treatment Arm",
       x = "Total Malaria Episodes",
       y = "Frequency",
       fill = "Treatment Arm") +
  theme_minimal()

*Bar Graph of Total Malaria Episodes During Pregnancy by Treatment Arm*

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(here)

# Recode 'total_malaria_episodes_during_pregnancy' into a categorical variable
promo_data_clean <- promo_data_clean %>%
  mutate(
    MalariaEpisodesPreg_cat = case_when(
      total_malaria_episodes_during_pregnancy %in% c(0, 1) ~ "1",
      total_malaria_episodes_during_pregnancy %in% c(2, 3) ~ "2–3",
      total_malaria_episodes_during_pregnancy >= 4 ~ "≥4"
    ),
    MalariaEpisodesPreg_cat = factor(MalariaEpisodesPreg_cat, levels = c("1", "2–3", "≥4"))
  )

# Create the bar graph
ggplot(promo_data_clean, aes(x = MalariaEpisodesPreg_cat, fill = study_arm)) +
  geom_bar(position = "dodge") +
  labs(title = "Total Malaria Episodes During Pregnancy by Treatment Arm",
       x = "Total Malaria Episodes During Pregnancy (Categorical)",
       y = "Count",
       fill = "Treatment Arm") +
  theme_minimal()

*Boxplot of Birthweight by IPTp Treatment Arm*

```{r}
ggplot(promo_data_clean, aes(x = study_arm, y = birth_weight, fill = study_arm)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Birthweight by IPTp Treatment Arm",
       x = "Treatment Arm",
       y = "Birthweight (kg)") +
  theme_minimal() +
  theme(legend.position = "none")

*Bar Graphs for Gravidity and Parity*

```{r}
# Load necessary libraries

# Recode Gravidity into categories: "1", "2–3", "≥4"
promo_data_clean <- promo_data_clean %>%
  mutate(
    Gravidity_cat = case_when(
      gravidity == 1 ~ "1",
      gravidity %in% c(2, 3) ~ "2–3",
      gravidity >= 4 ~ "≥4"
    ),
    Gravidity_cat = factor(Gravidity_cat, levels = c("1", "2–3", "≥4"))
  )

# Create the bar graph for Gravidity by Treatment Arm
ggplot(promo_data_clean, aes(x = Gravidity_cat, fill = study_arm)) +
  geom_bar(position = "dodge") +
  labs(title = "Gravidity Distribution by Treatment Arm",
       x = "Gravidity Category (1, 2–3, ≥4)",
       y = "Count",
       fill = "Treatment Arm") +
  theme_minimal()

*Parity Distribution by Treatment Arm*

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(here)

# Recode Parity into categories: "0", "1–2", "≥3"
promo_data_clean <- promo_data_clean %>%
  mutate(
    Parity_cat = case_when(
      parity == 0 ~ "0",
      parity %in% c(1, 2) ~ "1–2",
      parity >= 3 ~ "≥3"
    ),
    Parity_cat = factor(Parity_cat, levels = c("0", "1–2", "≥3"))
  )

# Create the bar graph for Parity by Treatment Arm
ggplot(promo_data_clean, aes(x = Parity_cat, fill = study_arm)) +
  geom_bar(position = "dodge") +
  labs(title = "Parity Distribution by Treatment Arm",
       x = "Parity Category (0, 1–2, ≥3)",
       y = "Count",
       fill = "Treatment Arm") +
  theme_minimal()

*Figure 1: Differential Impact of IPTp Treatment on the Relationship Between Malaria Episodes and Adverse Birth Outcomes*

```{r}
# Create the composite adverse outcome variable
promo_data_clean <- promo_data_clean %>%
  mutate(
    low_birth_weight = ifelse(birth_weight < 2.5, 1, 0),
    adverse_birth_outcome = ifelse(preterm_births_count > 0 | stillbirth_bin == 1 | low_birth_weight == 1, 1, 0)
  )

# Fit the interaction model (ensuring that adverse_birth_outcome now exists)
model_interaction <- glm(adverse_birth_outcome ~ total_malaria_episodes * study_arm + 
                           age_at_enrollment_years + gravidity + education_level,
                         family = binomial(link = "logit"),
                         data = promo_data_clean)

# Generate predicted probabilities over the range of total malaria episodes by study arm
pred <- ggeffect(model_interaction, terms = c("total_malaria_episodes [all]", "study_arm"))

# Plot the predicted probabilities
interaction_plot <- ggplot(pred, aes(x = x, y = predicted, color = group)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.2, color = NA) +
  labs(title = "Differential Impact of IPTp Treatment on the Relationship\nBetween Malaria Episodes and Adverse Birth Outcomes",
       x = "Total Malaria Episodes",
       y = "Predicted Probability of Adverse Outcome",
       color = "Treatment Arm",
       fill = "Treatment Arm") +
  theme_minimal()

interaction_plot

*Research questions 1& 2*

```{r}
# Convert date columns to Date objects
promo_data_clean <- promo_data_clean %>%
  mutate(
    enrollment_date = as.Date(enrollment_date, format = "%Y-%m-%d"),
    withdrawal_date = as.Date(withdrawal_date, format = "%Y-%m-%d"),
    child_withdrawal_date = as.Date(child_withdrawal_date, format = "%Y-%m-%d")
  )

@@

# Ensure the 'data/clean' directory exists
if (!dir.exists(here("data", "clean"))) {
  dir.create(here("data", "clean"), recursive = TRUE)
}

@@

```{r}
# Save the cleaned data (if not already saved)
write.csv(promo_data_clean, here::here("data", "clean", "PROMO_Data_clean.csv"), row.names = FALSE)

# Read in the cleaned data
promo_data_clean <- read.csv(here::here("data", "clean", "PROMO_Data_clean.csv"))

# Load necessary libraries
library(dplyr)
library(ggplot2)
library(broom)

### Data Preparation

# 1. Create a composite adverse birth outcome variable
#    Adverse outcome is defined as: 
#    - Preterm birth: preterm_births_count > 0 
#    OR stillbirth: stillbirth_bin == 1 
#    OR low birth weight: birth_weight < 2.5 kg
promo_data_clean <- promo_data_clean %>%
  mutate(low_birth_weight = ifelse(birth_weight < 2.5, 1, 0),
         adverse_birth_outcome = ifelse(preterm_births_count > 0 | stillbirth_bin == 1 | low_birth_weight == 1, 1, 0))

# 2. Create a subgroup variable for maternal age (<25 years vs. >=25 years)
promo_data_clean <- promo_data_clean %>%
  mutate(age_group = ifelse(age_at_enrollment_years < 25, "Young", "Older"))

### Research Question 1: 
### "Does the type of IPTp regimen modify the association between malaria episode frequency and adverse birth outcomes in Ugandan pregnant women?"

# Fit a logistic regression model with an interaction between total malaria episodes and study_arm
model_interaction <- glm(adverse_birth_outcome ~ total_malaria_episodes * study_arm + 
                           age_at_enrollment_years + gravidity + education_level,
                         family = binomial(link = "logit"),
                         data = promo_data_clean)

# Summarize and tidy the model output (exponentiating coefficients to yield odds ratios)
summary(model_interaction)
tidy_model_interaction <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)
print(tidy_model_interaction)

### Research Question 2: 
### "Among younger Ugandan pregnant women, is increased gravidity associated with a reduced risk of adverse birth outcomes?"

# Subset the data for young mothers (age < 25)
young_mothers <- promo_data_clean %>% filter(age_group == "Young")

# Fit a logistic regression model within the young mothers subgroup 
# to examine the association between gravidity and adverse birth outcomes
model_gravidity_young <- glm(adverse_birth_outcome ~ gravidity + total_malaria_episodes + education_level,
                             family = binomial(link = "logit"),
                             data = young_mothers)

# Summarize and tidy the subgroup model output
summary(model_gravidity_young)
tidy_model_gravidity_young <- tidy(model_gravidity_young, exponentiate = TRUE, conf.int = TRUE)
print(tidy_model_gravidity_young)

### Additional Exploratory Plots (Optional)

# Plot the distribution of total malaria episodes by study arm
ggplot(promo_data_clean, aes(x = study_arm, y = total_malaria_episodes)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Malaria Episode Frequency by Treatment Arm", 
       x = "Treatment Arm", y = "Total Malaria Episodes")

# Plot the proportion of adverse outcomes by gravidity among young mothers
ggplot(young_mothers, aes(x = gravidity, fill = factor(adverse_birth_outcome))) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Adverse Outcomes by Gravidity (Young Mothers)",
       x = "Gravidity", y = "Proportion", fill = "Adverse Outcome")

@@

##################################################################################################################################################################

*Question 1: "Does the type of IPTp regimen modify the association between malaria episode frequency and adverse birth outcomes in Ugandan pregnant women?"*

@@

# Create the composite adverse outcome variable:
# adverse_birth_outcome = 1 if any of the following occur:
# preterm birth (preterm_births_count > 0), stillbirth (stillbirth_bin == 1), or low birth weight (<2.5 kg)
promo_data_clean <- promo_data_clean %>%
  mutate(
    low_birth_weight = ifelse(birth_weight < 2.5, 1, 0),
    adverse_birth_outcome = ifelse(preterm_births_count > 0 | stillbirth_bin == 1 | low_birth_weight == 1, 1, 0)
  )

# Fit the logistic regression model with an interaction term
model_interaction <- glm(adverse_birth_outcome ~ total_malaria_episodes * study_arm +
                           age_at_enrollment_years + gravidity + education_level,
                         family = binomial(link = "logit"),
                         data = promo_data_clean)

# Display the model summary
summary(model_interaction)

# Tidy the model output (exponentiating coefficients to yield odds ratios)
tidy_model <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)
print(tidy_model)

# Generate predicted probabilities over the range of total malaria episodes by treatment arm
pred <- ggeffect(model_interaction, terms = c("total_malaria_episodes [all]", "study_arm"))

# Create the interaction plot
interaction_plot <- ggplot(pred, aes(x = x, y = predicted, color = group)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.2, color = NA) +
  labs(title = "Differential Impact of IPTp Treatment on the Relationship\nBetween Malaria Episodes and Adverse Birth Outcomes",
       x = "Total Malaria Episodes",
       y = "Predicted Probability of Adverse Outcome",
       color = "Treatment Arm",
       fill = "Treatment Arm") +
  theme_minimal()

interaction_plot

@@

*Table 3: Interaction Between Malaria Exposure and IPTp Treatment Arm in Predicting Adverse Birth Outcomes*

```{r}
# Tidy the model output with exponentiated coefficients (odds ratios)
tidy_model <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)

# Remove the intercept row and recode variable names using case_when()
tidy_model_filtered <- tidy_model %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      term == "total_malaria_episodes" ~ "Total Malaria Episodes",
      term == "study_armSP" ~ "Treatment Arm (SP)",
      term == "age_at_enrollment_years" ~ "Age at Enrollment (years)",
      term == "gravidity" ~ "Gravidity",
      term == "education_levelSecondary" ~ "Secondary Education",
      term == "education_levelTertiary" ~ "Tertiary Education",
      term == "education_levelUniversity" ~ "University Education",
      term == "total_malaria_episodes:study_armSP" ~ "Interaction: Malaria Episodes × SP",
      TRUE ~ term
    )
  )

# Round key values for clarity
tidy_model_filtered <- tidy_model_filtered %>%
  mutate(
    estimate = round(estimate, 2),
    std.error = round(std.error, 2),
    statistic = round(statistic, 2),
    p.value = round(p.value, 3),
    conf.low = round(conf.low, 2),
    conf.high = round(conf.high, 2)
  )

# Render the table with a blank header for the first column
kable(
  tidy_model_filtered,
  format = "pandoc",
  caption = "*Table X: Regression Results with VIFs*",
  col.names = c("", "Odds Ratio", "Std. Error", "z value", "p-value", "95% CI Lower", "95% CI Upper")
) %>% 
  kable_styling(full_width = FALSE, position = "center")

@@

*Fit a logistic regression model without the interaction term*

```{r}
# Fit a logistic regression model without the interaction term
model_no_interaction <- glm(adverse_birth_outcome ~ total_malaria_episodes + study_arm +
                              age_at_enrollment_years + gravidity + education_level,
                            family = binomial(link = "logit"),
                            data = promo_data_clean)

# Perform a Likelihood Ratio Test comparing models with and without the interaction
lrt <- anova(model_no_interaction, model_interaction, test = "LRT")
print(lrt)

# Check for multicollinearity using the VIF (Variance Inflation Factor)
library(car)
vif_values <- vif(model_interaction)
print(vif_values)

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
library(here)

# Generate Analysis of Deviance Table (Likelihood Ratio Test)
anova(model_no_interaction, model_interaction, test = "LRT")
anova_df <- as.data.frame(anova_out)
anova_df$Model <- c("Model 1 (No Interaction)", "Model 2 (With Interaction)")
anova_df <- anova_df %>% 
  select(Model, `Resid. Df`, `Resid. Dev`, Df, Deviance, `Pr(>Chi)`)
colnames(anova_df) <- c("Model", "Residual Df", "Residual Dev", "Df", "Deviance", "p-value")

kable(
  anova_df,
  format = "pandoc",
  caption = "*Table X: Analysis of Deviance Comparing Models With and Without Interaction*"
) %>% 
  kable_styling(full_width = FALSE, position = "center")

@@

```{r}
library(dplyr)
library(knitr)
library(kableExtra)
library(car)
library(here)

# Compute Variance Inflation Factors (VIF) for Model 2 (with interaction)
vif_vals <- vif(model2)
vif_df <- as.data.frame(vif_vals)
vif_df$Variable <- rownames(vif_vals)
vif_df <- vif_df %>% rename(`Df (VIF)` = Df)
vif_df <- vif_df %>% 
  select(Variable, GVIF, `Df (VIF)`, `GVIF^(1/(2*Df))`) %>%
  mutate(
    Variable = case_when(
      Variable == "total_malaria_episodes" ~ "Total Malaria Episodes",
      Variable == "study_arm" ~ "Treatment Arm",
      Variable == "age_at_enrollment_years" ~ "Age at Enrollment",
      Variable == "gravidity" ~ "Gravidity",
      Variable == "education_level" ~ "Education Level",
      Variable == "total_malaria_episodes:study_arm" ~ "Interaction: Malaria Episodes × SP",
      TRUE ~ Variable
    )
  )

kable(
  vif_df,
  format = "pandoc",
  caption = "*Table Y: Variance Inflation Factors for Model 2 (With Interaction)*"
) %>% 
  kable_styling(full_width = FALSE, position = "center")

@@

##################################################################################################################################################################

## Research Question 2:

*"Among younger (women under 25 years old) Ugandan pregnant women, is increased gravidity associated with a reduced risk of adverse birth outcomes?"*

```{r}
# Subset the data to include only women under 25 years of age
promo_data_young <- promo_data_clean %>%
  filter(age_at_enrollment_years < 25)

# Create the composite adverse birth outcome variable if not already done
promo_data_young <- promo_data_young %>%
  mutate(
    low_birth_weight = ifelse(birth_weight < 2.5, 1, 0),
    adverse_birth_outcome = ifelse(preterm_births_count > 0 | stillbirth_bin == 1 | low_birth_weight == 1, 1, 0)
  )
